Chess EDA

### 2023-03-17 owen/01_eda.qmd
some initial models

library(tidyverse)
library(RcppRoll)

@@

```{r}
a <- head(sort(table(c(bullet_60$White, bullet_60$Black)), decreasing = TRUE), 25)

@@

tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 5)

@@

>>>>

library(tidyverse)
library(RcppRoll)
library(cmdstanr)
library(bayesplot)

@@

```{r}
## want to use focal players not all players to get these lists to be the same
a <- head(sort(table(c(bullet_60$Username)), decreasing = TRUE), 17)
## if don't use all focal players here then removing data because they won't be
## involved

@@

tidy_games <- map_dfr(top_players, get_hist, bullet_60, prev_n = 10) %>% 
  as_tibble()

@@

Now need to figure out how to feed this into a Stan model.

```{r}
head(tidy_games)

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% mutate(ave_prop = focal_win_prop - mean(focal_win_prop))

init_data

### will need
### number of focal players
### which focal player is in each row
### their outcome 0 or 1
### their colour white (1) or black (0)
### ranking diff

```{r}

stan_data <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$focal_win_prop)

stan_file <- "model1.stan"

mod <- cmdstan_model(stan_file)

fit1 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit1$summary()

## but seems to give positive winner effects for all players
mcmc_hist(fit1$draws("alpha"))
mcmc_hist(fit1$draws("beta"))

# fit$save_object(file = "simple_multi_logistic.RDS")

Instead account for the covariates of colour and elo difference also.

```{r}
stan_file <- "model2.stan"

mod <- cmdstan_model(stan_file)

fit2 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit2$summary()

mcmc_hist(fit2$draws(c("beta", "gamma1", "gamma2")))

# fit$save_object(file = "cov_multi_logistic.RDS")

## these indicate clear positive coefficients for the betas,
## which are basically winner effects, account for the role
## of win history
## elo seems to have a much smaller role

## repeat for a different history, did this using past 5 games, could try
## others.
## could also compare to longer games, see if you get the same thing
## don't put a common prior on the betas, which is what the Cribben paper
## did. also don't standardize them

We can then instead include partial pooling of the $\beta$ coefficients.

```{r}
stan_file <- "model3.stan"

mod <- cmdstan_model(stan_file)

fit3 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3$summary()
mcmc_hist(fit3$draws(c("beta", "mu", "tau")))

mcmc_trace(fit3$draws(c("mu", "tau", "beta")))

## Repeat using normalized win Ratio

```{r}

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model1.stan"

mod <- cmdstan_model(stan_file)

fit1_ave <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit1_ave$summary()

## but seems to give positive winner effects for all players
mcmc_hist(fit1_ave$draws("alpha"))
mcmc_hist(fit1_ave$draws("beta"))

```{r}
stan_file <- "model2.stan"

mod <- cmdstan_model(stan_file)

fit2_ave <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit2_ave$summary()

mcmc_hist(fit2$draws(c("beta", "gamma1", "gamma2")))
```

```{r}
stan_file <- "model3.stan"

mod <- cmdstan_model(stan_file)

fit3_ave <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3_ave$summary()

mcmc_hist(fit3$draws(c("beta", "mu", "tau", "gamma1", "gamma2")))
```

Given this output, we can get the probability a player who is black
will win against and equally ranked player, when his current win ratio 
is exactly his average will be $1-\exp(-\alpha_j)$ while if they 
are playing as white their win probability will be
$1-\exp(-(\alpha_j + \gamma_1))$, so $\gamma_1$ being positive indicates the win
probability increases when playing white, which makes sense.
Playing white makes a player who has 50% chance of winning as black
have 54% chance of winning with white. Wikipedia says the difference is
4-6% so similar enough.
